Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-02 paper.tex
Wrote in Section IV (Experiments)

% Problem definition
Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. Big~Brain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. This dataset
is being re-scanned at an even higher resolution, leading to 100~TB of
raw data. Other examples found in medical imaging, our primary domain
of interest, include high-resolution 3D electron microscopy (see,
e.g., ~\cite{bock2011network}) or micro- and
nano-tomography~\cite{10.1371/journal.pone.0035691}. As such images
would typically be processed on a computing cluster, possibly using
locality-aware file systems such as HDFS~\cite{shvachko2010hadoop},
efficient software libraries are needed to split and merge them
efficiently. In particular, data reads and writes might result in
extensive seek times that drastically limit I/O performance.  In this
paper we introduce and compare a family of sequential algorithms to
split and merge images with minimal seek time.

@@

We consider a file format where voxels are written in column-major
order, that is, for an image of dimensions (i, j, k), i is the fastest
changing dimension, followed by dimension j, followed by dimension
k. All voxels in a \emph{slice} have a the same $k$ and all voxels in
a \emph{row} have the same $j$.  NIfTI, the file format defined by the
Neuroimaging Informatics Technology Initiative
NIfTI\footnote{\url{https://nifti.nimh.nih.gov}} is such a format,
used by the vast majority of the neuroimaging community.

@@

\begin{figure}
  \centering
  \begin{minipage}[b]{0.42\columnwidth}
    \def\svgwidth{0.8\columnwidth}
    \input{figures/svg/Notations.pdf_tex}
    \caption{Notations. A \emph{block row} is shown in red. A
      \emph{block slice} is shown in blue.}
    \label{fig:notations}
  \end{minipage}
  \quad \quad \quad
  \begin{minipage}[b]{0.42\columnwidth}
    \def\svgwidth{0.8\columnwidth}
    \input{figures/svg/buffer.pdf_tex}
    \caption{Buffer used in cluster reads (d=4).  White portions in
      the buffer are not allocated. }
    \label{fig:cluster-reads-buffer}
  \end{minipage}
\end{figure}
Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks} show
the naive merging methods for slices and blocks. These algorithms
actually have very different complexities even though blocks and
slices have identical sizes. Since slices are stored contiguously in
the reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$ as $n$ seeks are
required to read the slices and $n$ seeks are required to write them:
\begin{equation}
  N_\mathrm{slices} = 2n 
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block:
\begin{equation*}
N_\mathrm{blocks} = n+nd^2 
\end{equation*}
or, using $R$ and $n$ as main variables:
\begin{equation}
N_\mathrm{blocks} = n+n\left(\sqrt[3]{\frac{R}{n}}\right)^2
\end{equation}
In practice, this difference could lead to a tremendous slowdown, as
we will show later.

@@

Algorithm~\ref{algo:naive-slices} is a particular case of memory
buffering where the amount of available memory equals the maximum size
of a chunk. More buffering can be achieved when the amount of
available memory increases, as shown in
Algorithm~\ref{algo:buffered-slices}.
\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE sorted\_slices = sort slices by increasing k values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = sorted\_slices[i]
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}
This algorithm writes in the reconstructed image using a single seek
per memory load. Therefore:
\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}}
\end{equation}
Buffered slices are straightforward to implement, however, their
extension to block merging is not easy. The remainder of this Section
presents our attempts for such a generalization.

@@

\subsection{Analysis}

Based on a few simulations of Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads}, we conjectured that the comparison
between Cluster reads and Multiple reads essentially depends on the
three Cluster reads cases. When Cluster reads merge incomplete block
rows (case 1), they are always outperformed by Multiple reads
(\textbf{prop 1}).  When Cluster reads merge complete block rows (case
2), the comparison between Cluster reads and Multiple reads depends on
the values of $D$ and $n$ (\textbf{prop 2}). When Cluster reads merge complete
block slices (case 3), Cluster reads and Multiple reads are equivalent
(\textbf{prop 3}).

@@

The 5 algorithms presented earlier are implemented in a Python library
called \texttt{sam} (for ``split and merge''), available at
\url{https://github.com/big-data-lab-team/-bin-sam} under MIT
license. It uses Nibabel~\cite{matthew_brett_2016_60808} for image I/O
and NumPy for array manipulations.

@@

\section{Experiments}
\label{sec:results}

\subsection{Data}
To benchmark our library we use the 76~GB Big~Brain image split in 125
non-overlapping chunks. The data is available \todo{[here]}.

@@

We run experiments using a Dell Precision Tower 3620 with CentOS Linux release
7.3.1611, 32~GB of RAM and two disks:
\begin{itemize}
  \item Hard disk drive (HDD): HGST Travelstar 7K1000, 7200~rpm, 931GiB (1TB), firmware version JB0OA3W0.
    \item Solid-state drive (SSD): SanDisk X400 2.5, 238GiB (256GB),
      firmware version X4130012.
\end{itemize}
Both drives used 512-byte logical sectors, 4096-byte physical sectors,
SATA >3.1 (6.0 Gb/s) and were accessed through the XFS file system
v4.5.0.
%smartctl --xall /dev/sda

@@

We used release xxx of our library.

Experiment scripts are available at xxx.
As our experiments results are heavily rely on IO performance of the disk,
we drop the system cache to get more precise result before each experiment.

\subsection{Slices vs Blocks}

Figure~\ref{fig:slices-vs-blocks} shows the merging time for Slices vs
Blocks, on SSD and HDD. Merging with Blocks takes about twice the time
of Slices, which motivates the need for improved algorithms to merge
blocks. As expected, seek time is mainly responsible for this
difference, although differences in read times are also observed,
perhaps due to a caching effect at the disk or OS level. This behavior is
reproduced in SSD and HDD, which shows that SSDs might not be the only
solution to reduce seek time.  \todo{On figure: check that breakdowns
  add up to total time. Harmonize time formats. Harmonize colors (all
  slices in purple, all blocks in green). If time permits, try
  histogram for breakdown, on same figure than total time.}

@@

\subsection{Cluster reads}

@@

\subsection{Multiple reads}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{figures/number-of-seeks.pdf}
  \hfill
  \caption{Number of seeks for all algorithms. y scale is logarithmic. Each algorithm is
    represented with a different color. Dark color is experimental
    value; bright color is model.}
  % Gnuplot file: scripts/number_of_seeks.gnuplot
\label{fig:number-of-seeks}
\end{figure}

Results discussion:
\begin{itemize}
\item Comment on calculation time.
\item Write time is higher than read time because use nibabel for writing but not reading.
\item It seems that the merge time is quite stable w.r.t the amount of available memory. This is good news for multiple reads. 
\end{itemize}

\subsection{Cluster reads vs Multiple reads}

\subsection{Lossless Compression}

@@

\subsubsection{Multiple writes}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-ssd.pdf}
  \hfill
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-hdd.pdf}
  \caption{Comparison between Multiple writes, blocks and slices. Left: SSD. Right: HDD.}
  % Gnuplot file: scripts/blocks-slices/createboxplot.gnplt
\label{fig:multiple-writes}
\end{figure}

@@

@@

% Library is available. We recommend using xxx for most cases. Same
% algo works for blocks or slices.

>>>>

% Problem definition
Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. Big~Brain, for
\todo{summarize sizes} instance, is a 3D histological image of the human brain that
represents 1~TB of raw data organized in 3600 planes at full resolution
and 76~GB at a 40-micrometer isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. This dataset
is being re-scanned at an even higher resolution, leading to 100~TB of
raw data. Other examples found in medical imaging, our primary domain
of interest, include high-resolution 3D electron microscopy (see,
e.g., ~\cite{bock2011network}) or micro- and
nano-tomography~\cite{10.1371/journal.pone.0035691}. As such images
would typically be processed on a computing cluster, possibly using
locality-aware file systems such as HDFS~\cite{shvachko2010hadoop},
efficient software libraries are needed to split and merge them
efficiently. In particular, data reads and writes might result in
extensive seek times that drastically limit I/O performance.  In this
paper we introduce and compare a family of sequential algorithms to
split and merge images with minimal seek time.

@@

We consider a file format where voxels are written in column-major
order, that is, for an image of dimensions (i, j, k), i is the fastest
changing dimension, followed by dimension j, followed by dimension
k. All voxels in a \emph{slice} have a the same $k$ and all voxels in
a \emph{row} have the same $j$.  NIfTI, the file format defined by the
Neuroimaging Informatics Technology Initiative
NIfTI (\url{https://nifti.nimh.nih.gov}) is such a format,
used by the vast majority of the neuroimaging community.

@@

\begin{figure}
  \centering
  \begin{minipage}[b]{0.42\columnwidth}
    \def\svgwidth{0.8\columnwidth}
    \input{figures/svg/Notations.pdf_tex}
    \caption{Notations. A \emph{block row} is shown in red. A
      \emph{block slice} is shown in blue.}
    \label{fig:notations}
  \end{minipage}
  \quad \quad \quad
  \begin{minipage}[b]{0.42\columnwidth}
    \def\svgwidth{0.8\columnwidth}
    \input{figures/svg/buffer.pdf_tex}
    \caption{Buffer used in cluster reads (d=4).  White portions in
      the buffer are not allocated. }
    \label{fig:cluster-reads-buffer}
  \end{minipage}
\end{figure}
Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks} show
the naive merging methods for slices and blocks. These algorithms
actually have very different complexities even though blocks and
slices have identical sizes. Since slices are stored contiguously in
the reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$ as $n$ seeks are
required to read the slices and $n$ seeks are required to write them:
\begin{equation}
  N_\mathrm{slices} = 2n \label{eq:naive-blocks}
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block:
\begin{equation*}
N_\mathrm{blocks} = n+nd^2  
\end{equation*}
or, using $R$ and $n$ as main variables:
\begin{equation}
N_\mathrm{blocks} = n+n\left(\sqrt[3]{\frac{R}{n}}\right)^2 \label{eq:naive-slices}
\end{equation}
In practice, this difference could lead to a tremendous slowdown, as
we will show later.

@@

Algorithm~\ref{algo:naive-slices} is a particular case of memory
buffering where the amount of available memory equals the maximum size
of a chunk. More buffering can be achieved when the amount of
available memory increases, as shown in
Algorithm~\ref{algo:buffered-slices}.
\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE sorted\_slices = sort slices by increasing k values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = sorted\_slices[i]
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}
This algorithm writes in the reconstructed image using a single seek
per memory load. Therefore:
\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}} \label{eq:buff-slices}
\end{equation}
Buffered slices are straightforward to implement, however, their
extension to block merging is not easy. The remainder of this Section
presents our attempts for such a generalization.

@@

\subsection{Analysis}
\label{sec:analysis}
Based on a few simulations of Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads}, we conjectured that the comparison
between Cluster reads and Multiple reads essentially depends on the
three Cluster reads cases. When Cluster reads merge incomplete block
rows (case 1), they are always outperformed by Multiple reads
(\textbf{prop 1}).  When Cluster reads merge complete block rows (case
2), the comparison between Cluster reads and Multiple reads depends on
the values of $D$ and $n$ (\textbf{prop 2}). When Cluster reads merge complete
block slices (case 3), Cluster reads and Multiple reads are equivalent
(\textbf{prop 3}).

@@

The 5 algorithms presented earlier are implemented in a Python library
called \texttt{sam} (for ``split and merge''). It uses Nibabel~\cite{matthew_brett_2016_60808} for image I/O
and NumPy for array manipulations.

@@

\section{Experiments}
\label{sec:experiments}

\subsection{Data}
We used the 3850x3025x3500 Big~Brain image split in 125
non-overlapping chunks of size 770x605x700 with 2 bytes per voxel
(total size, uncompressed is 75.92~GB). Big~Brain is a reference brain
based on the reconstruction of 7404 histological sections at nearly
cellular resolution of 20 micrometers~\cite{amunts2013bigbrain}. It is
a freely, publicly available tool with numerous applications in
neurosciences and neurosurgery.

We used the blocks of the 2015 Big~Brain release with 40-micrometer
isotropic resolution available at
\url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D\_Blocks/40um}. We
converted them to NiFTI 1.0 using Nibabel and left them uncompressed.

\todo{add slices}

@@

We used a Dell Precision Tower 3620 workstation with CentOS Linux
release 7.3.1611, 32~GB of RAM and two disks. (1) a Hard disk drive
(HDD): HGST Travelstar 7K1000, 7200~rpm, 931GiB (1TB), firmware
version JB0OA3W0; (2) a Solid-state drive (SSD): SanDisk X400 2.5,
238GiB (256GB), firmware version X4130012.  Both drives used 512-byte
logical sectors, 4096-byte physical sectors, SATA \textgreater 3.1 (6.0 Gb/s) and
were accessed through the XFS file system v4.5.0. We used
\texttt{iotop} (\url{http://guichaz.free.fr/iotop}) to monitor I/Os on
the workstation and make sure that no other process was compromising
our measures.
%smartctl --xall /dev/sda

@@

We used Git tag 0.1 of our \texttt{sam} library. Our experiment
scripts are available at
\url{https://github.com/big-data-lab-team/paper-sequential-split-merge/blob/master/scripts/experiment}. We
used them to split and merge using Buffered slices, Cluster reads and
Multiple reads, with 3GB, 6GB, 9GB, 12GB and 16GB of memory. We also
did 0GB of memory for Buffered slices and Cluster reads, which
triggered naive slice and block merging. We did 5 repetitions for each
memory value. Memory values were shuffled in each repetition, to
correct for potential biases coming from ordering, such as caching
effects. To improve reproducibility, we dropped the kernel page,
dentry and inode caches before each run (\texttt{echo 3 | sudo tee
  /proc/sys/vm/drop\_caches}).

\subsection{Results}

\subsubsection{Merging time}

% For each paragraph below, comment on overal trend and breakdown.

% Slices vs blocks
Figure~\ref{fig:slices-vs-blocks} shows the merging time for Slices vs
Blocks, on SSD and HDD. Merging with Blocks takes about twice the time
of Slices, which motivates the need for improved algorithms to merge
blocks. As expected, seek time is mainly responsible for this
difference, although differences in read times are also observed,
perhaps due to a caching effect at the disk or OS level. This behavior is
reproduced in SSD and HDD, which shows that SSDs might not be the only
solution to reduce seek time.  \todo{On figure: check that breakdowns
  add up to total time. Harmonize time formats. Harmonize colors (all
  slices in purple, all blocks in green). If time permits, try
  histogram for breakdown, on same figure than total time.}

% Buffered slices

% Cluster reads
%% ultimately fine but not stable due to discontinuities
At 3GB, Cluster reads were in case 1, which explains that more than 10
million seeks were done, almost equivalent to naive blocks. At 6, 9
and 12~GB, Cluster reads were in case 2 and at 16~GB they were in case
3. This confirms that Cluster reads are beneficial in case 2 and 3
only, consistently with the analysis in Section~\ref{sec:analysis} .

% Multiple reads
%% more stable

\begin{itemize}
\item Comment on calculation time.
\item Write time is higher than read time because use nibabel for writing but not reading.
\item It seems that the merge time is quite stable w.r.t the amount of available memory. This is good news for multiple reads. 
\end{itemize}

@@

@@

@@

\subsubsection{Seeks}

The number of seeks is reported in Figure~\ref{fig:number-of-seeks},
for all algorithms and the corresponding models
(Equations~\ref{eq:naive-blocks}
to~\ref{eq:seeks-multiple-reads}). Note the logarithmic y scale. Error
bars are not reported as numbers of seeks were constant across all
repetitions. The mean model error is \todo{x\%}, explained by the fact
that the model assumes cubic blocks while we used non-cubic ones in
the experiment. Overall, the model is validated as it correctly
explains the observations.   The measured seek times correlated with
the measured merging times ($\rho$=\todo{x}), which suggest that the
observed time reductions were indeed caused by reduced seek times.
\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{figures/number-of-seeks.pdf}
  \hfill
  \caption{Number of seeks for all algorithms. y scale is logarithmic. Each algorithm is
    represented with a different color. Dark color is experimental
    value; bright color is model.}
  % Gnuplot file: scripts/number_of_seeks.gnuplot
\label{fig:number-of-seeks}
\end{figure}

% correlated to merging time

% model error

@@

% Time

% Number of seeks

\begin{figure}[h]
  \centering
  \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-ssd.pdf}
  \hfill
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-hdd.pdf}
  \caption{Comparison between Multiple writes, blocks and slices. Left: SSD. Right: HDD.}
  % Gnuplot file: scripts/blocks-slices/createboxplot.gnplt
\label{fig:multiple-writes}
\end{figure}

@@

% Library is available. 
% overall, multiple reads is recommended as default parameters for the
% library. Cluster reads could be useful in some cases.

%Same algo works for blocks or slices.

@@

The complete Latex sources of this paper, including the collected data
and Gnuplot scripts used to generate the results figures are available
at \url{https://github.com/big-data-lab-team/paper-sequential-split-merge} under
GPLv3 license. Our \texttt{sam} library is at
\url{https://github.com/big-data-lab-team/sam}, under MIT license.
